Binning Strategies and Related Loss for Large Data

### 2015-04-10 Paper.Rnw
edits to section 6 intro

\ktm{
The exploration of loss of spatial and frequency information in the process of creating a binned scatterplot yielded a number of important properties and practical recommendations for their construction. Spatially binning with smaller bin dimensions will lead to lower spatial losses; however, finer binning requires more processing (Would like to use an O(n)-type argument, but I don't rightly understand this comp. sci. notation) and does not highlight large scale density structure. Ideally we would like to strike the optimal trade-off between bin size and spatial loss, but unfortunately the simulated data from the three bivariate distributions indicated that this relationship between them was nearly linear. Thus it is left to the plot designer to decide if they wish to highlight high level density structure with higher spatial loss or finer density structure with lower spatial loss.

For spatial binning, the random binning algorithm displayed strong advantage of avoiding the visual ``stripes'' problem that occurred when data recorded to a coarse resolution was binned using a bin width that was a non-integer multiple of the resolution, $\alpha$. However, standard binning algorithm is superior due to lower spatial loss, and data with coarse resolution is not problematic so long as a binwidth that is an integer multiple of $\alpha$ is selected.  It was also shown that in this scenario, we should use an origin that is offset to $\alpha / 2$ units below the minimum data value leads to minimize spatial loss.

If we elect to use frequency binning to aid the perceptual mapping tile shades to the plot key for frequency, it is recommended to use between four and seven distinct shades.  This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously. The standard and quantile algorithms for frequency binning have very similar loss properties for this range, so the decision falls to the interpretability of the resulting frequency bins. Log frequency binning is an alternative if the bin counts are heavily right skewed and it is desired to differentiate bins with low density.

These recommendations for binned scatterplot construction and loss properties will be demonstrated in the following examples using real bivariate data. First we will further explore the baseball pitching data.  Additionally we will investigate the relationship between scheduled and actual departure times for a full year of American commercial airline data. 
}

>>>>

The exploration of loss of spatial and frequency information in the process of creating a binned scatterplot yielded a number of important properties and practical recommendations for their construction. Spatially binning with smaller bin dimensions will lead to lower spatial losses; however, finer binning requires more processing time and does not highlight large scale density structure. Ideally we would like to strike the optimal trade-off between bin size and spatial loss, but unfortunately the simulated data from the three bivariate distributions indicated that this relationship between them was nearly linear. Thus it is left to the plot designer to decide if they wish to highlight high level density structure with higher spatial loss or finer density structure with lower spatial loss.

For spatial binning, the random binning algorithm displayed strong advantage of avoiding the visual ``stripes'' problem that occurred when data recorded to a coarse resolution was binned using a bin width that was a non-integer multiple of the resolution, $\alpha$. However, standard binning algorithm is superior due to lower spatial loss, and data with coarse resolution of $\alpha$ units can avoid the stripe illusion if a binwidth that is an integer multiple of $\alpha$ is selected.  It was also shown that in this scenario, a reasonable default for the binning origin is offset to $\alpha / 2$ units below the minimum data value because it will minimize spatial loss for symmetric data and generally preform well even for non-symmetric data.

If we elect to use frequency binning to aid the perceptual mapping tile shades to the plot key for frequency, it is recommended to use between four and seven distinct shades.  This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously. The standard and quantile algorithms for frequency binning have very similar loss properties for this range, so the decision becomes one of how we wish the shade of a tile to be interpreted. Standard log frequency binning is an alternative if the bin counts are heavily right skewed and it is desired to differentiate bins with low density. We also discovered that if we elect to use quantile binning, the log transformation will have no effect on the binned scatterplot shading due to the identical quantile-based mapping to ordinal frequency groupings for all monotone transformations on bin counts.

These recommendations for binned scatterplot construction and loss properties will be demonstrated in the following examples using real bivariate data. First we will further explore the baseball pitching data. Additionally we will investigate the relationship between scheduled and actual departure times for a full year of American commercial airline data.
